Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed particle hydrodynamics

### 2017-09-21 Manuscripts_GMD_ap_ref1.tex
9/20 edits ap

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades, ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more comprehensive physics effects. 
For example, FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient due to turbulence in the rising buoyant jet and the crosswind field. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is usually ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, there is need for greater skills in hazards forecasts especially where the plume model is used to generate source conditions for complex 2D (two dimensional) and 3D VATD models.

@@

To summarize, each 3D model has its own focus based on the problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation (depending on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depending on number of governing equations, resolution, numerical methods and parallel techniques) are always a pair of conflicting considerations in 3D plume simulations.

\subsection{Features of SPH}
To the best of our knowledge, all of the existing 3D plume models use mesh based Eulerian methods, and there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment;
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}

@@

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiples phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and computationally efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

@@

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step is used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible flow, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adapted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing. The core solver of our model is parallelized by MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) induced indexing and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors would play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not yet considered in this model. Introducing wind effects in our model requires dynamic pressure boundary conditions, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refered as a weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano on April 4th 1982 \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcanic plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

@@

\subsubsection{Open outlet pressure boundary condition}
The pressure of the surrounding atmosphere is given. Except for the pressure, boundary values for density, velocity, and energy on the outlet should depend on the solution. As we ignore the viscosity, the shear stress is ignored and normal stress (whose magnitude equals to pressure) balances the ambient pressure.
\begin{equation}
p = p_a\left(z\right)\label{eq:pressure_bc_p} 
\end{equation}

@@

\subsection{Artificial viscosity} \label{sec:artificial-viscosity}
In classical SPH, shock waves are handled by introducing artificial viscosity, a term that defined based on second derivatives of velocity, to smear out discontinuities. As in the case of first order derivatives, second order derivatives can be estimated by differentiating a SPH interpolation twice. However, such a formulation has two disadvantages: first, it is very sensitive to irregular distribution of particles, second, the second derivative of the kernel can change sign and lead to unphysical representations (for example, viscous dissipation causes decrease of the entropy).

@@

The SPH viscosity can be related to a continuum viscosity by converting the summation to integrals \citep{monaghan2005smoothed}. It has been shown that the shear viscosity coefficient $\eta = \frac{\rho \alpha h c}{8} $ and the bulk viscosity coefficient $ \zeta = \frac{5 \eta}{3}$ for two dimensional and $\eta = \frac{\rho \alpha h c}{10} $ , $ \zeta = \frac{5 \eta}{3}$ for three dimensional.
An extra term was added to $\nu$ considering aspects of the dissipative term in shock solutions based on Riemann solvers and lead to a new formulation of artificial viscosity. We adopt this new formulation in our simulation:
\begin{equation}
\Pi_{ab}^{\beta} = 
\begin{cases} 
      \dfrac{- \alpha \mu_{ab} \bar{c}_{ab} + \beta \mu_{ab}^2} {\bar{\rho}_{ab}} & \textbf{v}_{ab} \cdot \textbf{x}_{ab} < 0\\
      0 & \textbf{v}_{ab} \cdot \textbf{x}_{ab} > 0
\end{cases}
\label{eq:art-vis-shock}
\end{equation}
where
\begin{equation}
\mu_{ab} = \dfrac{h \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2} 
\end{equation}
%$\alpha$ and $\beta$ are two parameters that are free to be adjusted for each case. 
$\alpha$ and $\beta$ are two parameters that can be adjusted for different cases.
$\alpha = 1$ and $\beta = 2$ are  recommended by Monaghan for best results. In our simulation, these two parameters are calibrated to  $\alpha = 0.3$ and $\beta = 0.6$. $\eta$ is usually taken as 0.1 to prevent singularities.
%when $\textbf{x}_{ab} = 0$.

@@

We highlight an important feature of the SPH methodology. Adding new physics and new phases into the model is trivial in terms of discretization. For example, adding of new source (or sink) into Eq. (\ref{eq:ns-sph-d}), adding a drag force into Eq. (\ref{eq:ns-sph-v})  and adding a heat exchange term into Eq. (\ref{eq:ns-sph-e}) leads to the new discretization form:
\begin{align}
<\rho_a> = \sum m_b w_{ab} \left(h\right) + \dot{\rho}\left(\textbf{x},t\right)\label{eq:ns-source-sph-d} \\
\left\langle\dfrac{d \textbf{v}_a}{d t}\right\rangle= -\sum_b m_b \left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \nabla_a w_{a b}\left(h\right) +\textbf{g} + D \sum	_b m_b \dfrac{\textbf{v}_b - \textbf{v}_a}{\rho_b} \label{eq:ns-drag-sph-v} \\
\left\langle\dfrac{d e_a}{d t}\right\rangle=
 0.5\sum_b m_b \textbf{v}_{a b}\left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \cdot \nabla_a w_{a b}\left(h\right) + \sum_b \dfrac{m_b}{\rho_b}\left(\kappa_a + \kappa_b\right) \dfrac{\left(T_a - T_b\right)}{\textbf{r}_a - \textbf{r}_b} w_{ab}\left(h\right) \label{eq:ns-conduction-sph-e}
\end{align}
where the source term $\dot{\rho}$ can be a "sink" of erupted vapor due to its phase change.
%(see microphysics modelling in ATHAM \citep{oberhuber1998volcanic}).
$D$ is a drag force coefficient. $\kappa$ is the heat conduction coefficient. $T$ is the temperature. Other physics can be added easily in a similar way. Adding of these new terms leads to modification of several lines in the source code. The drag force term should show up when dynamics disequilibrium between different phases is considered. In that case, each phase needs one set of governing equations of Navier-Stokes type. Adding of new phase into SPH code only needs adding few new lines for the new phase besides interaction terms introduced by the new phase.

@@

\subsection{Tensile instability and corrected derivatives}
The classical SPH method was known to suffer from tensile instability and boundary deficiency. Tests of the standard SPH method indicate an instability in the tensile regime, while the calculations are stable in compression.  A simplest example of such test calculation exhibiting the instability involves a body which is subjected to an uniform initial stress, either compressive or tensile. If the initial stress is tensile, a very small velocity perturbation on a single particle can lead to particles clumping together, forming large voids and seriously corrupting density distribution. But if the initial stress is compressive, the small velocity perturbation on a single particle can not lead to any changes in particle distribution \citep{swegle1995smoothed}. To address these difficulties, \citep{chen1999improvement} proposed a corrected SPH formulation. For 1D case, employing a Taylor expansion for $A\left(x\right)$ about $x_a$, multiplying both sides by kernel function and then doing integration over the domain gives
\begin{equation}
\int_{\Omega} A\left(x\right) w\left(x- x_a, h\right) dx = 
A_a \int_{\Omega} w\left(x - x_a, h\right) dx +\dfrac{\partial A}{\partial x}(x_a) \int_{\Omega} \left(x-x_a\right) w\left(x - x_a, h\right) dx +...
\end{equation}
Ignoring derivative terms higher than first order, and writing the integral in particle approximation form leads to:
\begin{equation}
A_a = \frac{\sum_b m_b \dfrac{A_b}{\rho_b} w\left(x_a-x_b, h\right)}{\sum_b m_b \dfrac{1}{\rho_b} w\left(x_a-x_b, h\right)}
\label{eq:CSP-function-approximation-1d}
\end{equation}
Notice that the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) is actually summation approximation of Eq. (\ref{eq:SPH-kernel-normalization-prop}). That is to say, Eq. (\ref{eq:CSP-function-approximation-1d}) and Eq. (\ref{eq:SPH-approximation-sum}) are the same for particles far away from boundaries as the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) becomes $1$ in that case.
The first order derivative term can be obtained in a similar way:
\begin{equation}
\nabla A_a = \frac{\sum_b m_b \dfrac{A_b - A_a}{\rho_b} \nabla_a w\left(x_a-x_b, h\right)}{\sum_b m_b \dfrac{x_b - x_a}{\rho_b} \nabla_a w\left(x_a-x_b, h\right)}
\end{equation}

@@

\subsection{Boundary conditions}
All boundary conditions are imposed by ghost particles. Subfigure $b$ of Fig. \ref{fig:bc_and_domain_decomp} shows how boundaries are deployed. 
%Physical quantities of particles, including real particles and different types of ghost particles, are updated differently.
\subsubsection{Wall boundary condition}
Traditionally either ghost particles that mirror real particles across the boundary \citep {ferrari2009new} or boundary forces \citep {monaghan2009sph} have been used to impose the wall boundary conditions. One disadvantage of the latter  is that the boundary forces tend to corrupt the solution in the local neighborhood. In addition, a natural way of imposing eruption boundary condition is using eruption ghost particles. To impose boundary conditions in a consistent way, we adopt a modified version of the ghost particle method \citep {kumar2013parallel} for wall boundary conditions. Stationary wall ghost particles are deployed in the same way as real particles. Instead of enforcing symmetry particle by particle, a symmetric field across the boundary is explicitly enforced. Ghost particles are reflected into the domain and physical quantities are calculated at these reflected positions by SPH interpolations. It should be noted that wall ghost particles should not be counted when computing physical properties of these reflected positions. Assign all properties, except for velocity, at the corresponding reflected position to the ghost particle. The velocity of each wall ghost particles is set to have the same value but opposite direction of the interpolated velocity at its corresponding reflection. By this way, the no-slip wall boundary condition (Eq. (\ref{eq:wall_bc_v})) is imposed naturally. These wall ghost particles serve as neighbors in momentum and energy update. More implementation details about this method can be found in \citep {kumar2013parallel}. As these wall ghost particles are stationary, there is no mass flux on the boundary (Eq. (\ref{eq:wall_bc_rho}) and (\ref{eq:wall_bc_xi})). In addition, as temperature is also symmetric with respect to the boundary, the gradient of temperature vanishes and hence there is no internal energy flux on the wall boundary (Eq. (\ref{eq:wall_bc_e})). 
In our current model, the ground is assumed to be a flat ground. For more complicated topography, it has been shown in other paper \citep {kumar2013parallel} that it works as well. One of our concern regarding potential limitation of such method is deployment of ghost particles for more complicated boundaries which is not described in detail in the paper.
\subsubsection{Eruption boundary condition}
A natural way of imposing eruption boundary condition is using ghost particles that move with the eruption velocity and bear the temperature of the erupted material. A parabolic velocity profile that represents a fully developed Hagen-Poiseuille flow is used to determine the inlet particle velocity. The detailed shape of the parabolic profile is determined based on an averaged eruption velocity (Eq. (\ref{eq:erupt_bc_v})). The mass of eruption ghost particles is set to a value so that evaluation of Eq. (\ref{eq:ns-sph-d}) can return us a density that is consistent with the value given in Eq. (\ref{eq:erupt_bc_rho}). 
The internal energy associated with these particles are set to a value so that Eq. (\ref{eq:erupt_bc_e}) is satisfied. The mass fraction of erupted material (Eq. (\ref{eq:erupt_bc_xi})) is automatically satisfied as all particles in the eruption conduit are of phase 2. The density, momentum, and internal energy of these eruption ghost particles are not updated before they move above ground. As soon as they move out from erupt conduit, these ghost particles will be shifted to real particles and their physical quantities and position will be updated based on discretized governing equations. New ghost particles need to be added at the bottom of the eruption conduit as these existing ghost particles move upwards.

\subsubsection{Pressure boundary condition}
Another boundary condition in our model is the pressure outlet boundary. For flow in a straight channel, it is possible to treat the exit the same as the entry with a prescribed velocity profile. For flow with more complex channel, an exit far downstream of the flow disturbance is also feasible. However, the natural boundary condition (Eq. (\ref{eq:pressure_bc_p})) is more suitable for plume simulation as the outlet is open atmosphere. The way we impose pressure boundary condition is: adding several layers of pressure ghost particles surrounding the real atmosphere particles. Pressure, density and temperature are determined based on the elevation of particles. Velocity is set to zero for static atmosphere. The physical quantities for pressure ghost particles are not updated while these for real particles are updated at every time step. As position of all pressure ghost particles keep constant, we essentially impose a static pressure boundary condition. Real particles are removed as soon as they move out the pressure boundary.

>>>>

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades, ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more comprehensive physics effects. 
For example, FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient due to turbulence in the rising buoyant jet and the crosswind field. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is usually ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, there is need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex 2D (two dimensional) and 3D VATD models.

@@

To summarize, each 3D model has its own focus based on the problem of interest and modeling/numerical choices made.
 %Different simplification was made based on their scope. Improvements have been making since their birth. 
 Accuracy of simulation (depending on comprehensiveness of the model, resolution of discretization, numerical error, and order of accuracy) and simulation time (depending on number of governing equations, resolution, numerical methods and parallel techniques) are always conflicting considerations in 3D plume simulations.

\subsection{Features of SPH}
To the best of our knowledge, all of the existing 3D plume models use mesh based Eulerian methods, and there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. Currently it is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment;
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}

@@

As discussed in the previous paragraph, existing 3D plume models focus on one or several specific aspects of plume and have been extended to be more comprehensive by accounting for more physics or more phases. Easy extensibility and capability of handling multiple phase flow without additional numerical effort greatly facilitate future extension of SPH models. As volcanic plumes are in nature multiphase and without pre-defined boundary in the atmosphere, SPH is a suitable numerical method for plume modeling. The core physics, such as entrainment of air and thermal expansion, are essential for all plume modeling while some other physics, such as water condensation and aggregation etc., are important in specific scenarios. As an initial effort on exploiting advantages of SPH in volcanic plume modeling, we focus on capturing basic features in plume development using a robust numerically and computationally efficient framework with support for scalable parallel computing. %So our current model is far away from comprehensive.

@@

\begin{itemize}
\item We develop methodology to impose pressure boundary conditions by adding extra layers of static ghost particles. Additional constraints on the time step are used to avoid the growth of numerical fluctuations near the pressure boundary. We impose a velocity inlet boundary condition by placing several layers of ghost particles moving with eruption velocity. 
\item Turbulence model is crucial for reproducing the entrainment of air. There are several turbulence models proposed for SPH method \citep{issa2005numerical,violeau2007numerical}. We adopt a LANS (Lagrangian Averaged Navier-Stokes) turbulence model \citep{monaghan2011turbulence}, which was originally proposed for incompressible flow, and is extended here for compressible flow accounting for turbulent heat exchange. 
\item Corrected formulation of SPH \citep{chen1999improvement} is adopted to bypass the well-known tensile instability issues of classical SPH.
\item Simulation of volcanic plumes with acceptable accuracy requires fine resolution (very high particle counts) that cannot be accomplished without parallel computing large process counts. The core solver of our model is parallelized by distributed memory MPI (message passing interface standard) parallelism. In addition, a dynamic load balancing strategy is also developed. 
\item Imposition of some types of boundary conditions (such as eruption boundary condition) requires dynamically adding and removing of particles during simulation. To address this issue, we adopt an efficient data management scheme based on time dependent SFC (space filling curve) induced indexing and hash table. The computational cost is further reduced by adjusting simulation domain adaptively.
\end{itemize}

@@

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors would play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not yet considered in this model. Introducing wind effects in our model requires dynamic pressure boundary conditions, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refered as a weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as an eruption of \textit{El Chich{\'o}n} volcano on April 4th 1982 \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcanic plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

@@

\subsubsection{Open outlet pressure boundary condition}
The pressure of the surrounding atmosphere is given. Except for the pressure, boundary values for density, velocity, and energy on the outlet should depend on the solution. As we ignore the viscosity, the shear stress is ignored and normal stress (whose magnitude equals to pressure) balances the ambient pressure.
\begin{equation}
p = p_a\left(z\right)  \label{eq:pressure_bc_p} 
\end{equation}

@@

\subsection{Artificial Viscosity} \label{sec:artificial-viscosity}
In classical SPH, shock waves are handled by introducing artificial viscosity, a term that is defined based on second derivatives of velocity, to smear out discontinuities. As in the case of first order derivatives, second order derivatives can be estimated by differentiating a SPH interpolation twice. However, such a formulation has two disadvantages: first, it is very sensitive to irregular distribution of particles, second, the second derivative of the kernel can change sign and lead to unphysical representations (for example, viscous dissipation causes decrease of the entropy).

@@

The SPH viscosity can be related to a continuum viscosity by converting the summation to integrals \citep{monaghan2005smoothed}. It has been shown that  shear viscosity coefficient $\eta = \frac{\rho \alpha h c}{8} $ and bulk viscosity coefficient $ \zeta = \frac{5 \eta}{3}$ are appropriate for two dimensional flows and modified to $\eta = \frac{\rho \alpha h c}{10} $ , $ \zeta = \frac{5 \eta}{3}$ for three dimensional flows.
An extra term was added to $\nu$ considering aspects of the dissipative term in shock solutions based on Riemann solvers and lead to a new formulation of artificial viscosity. We adopt this new formulation in our simulation:
\begin{equation}
\Pi_{ab}^{\beta} = 
\begin{cases} 
      \dfrac{- \alpha \mu_{ab} \bar{c}_{ab} + \beta \mu_{ab}^2} {\bar{\rho}_{ab}} & \textbf{v}_{ab} \cdot \textbf{x}_{ab} < 0\\
      0 & \textbf{v}_{ab} \cdot \textbf{x}_{ab} > 0
\end{cases}
\label{eq:art-vis-shock}
\end{equation}
where
\begin{equation}
\mu_{ab} = \dfrac{h \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2} 
\end{equation}
%$\alpha$ and $\beta$ are two parameters that are free to be adjusted for each case. 
$\alpha$ and $\beta$ are two parameters that can be adjusted for different cases.
$\alpha = 1$ and $\beta = 2$ are  recommended by Monaghan for best results. In our simulation, these two parameters are calibrated to  $\alpha = 0.3$ and $\beta = 0.6$. $\eta$ is usually taken as 0.1 to prevent singularities.
%when $\textbf{x}_{ab} = 0$.

@@

We highlight an important feature of the SPH methodology. Adding new physics and new phases into the model is trivial in terms of discretization. For example, adding of new source (or sink) into Eq. (\ref{eq:ns-sph-d}), adding a drag force into Eq. (\ref{eq:ns-sph-v})  and adding a heat exchange term into Eq. (\ref{eq:ns-sph-e}) leads to the new discretization form:
\begin{align}
<\rho_a> = \sum m_b w_{ab} \left(h\right) + \dot{\rho}\left(\textbf{x},t\right)\label{eq:ns-source-sph-d} \\
\left\langle\dfrac{d \textbf{v}_a}{d t}\right\rangle= -\sum_b m_b \left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \nabla_a w_{a b}\left(h\right) +\textbf{g} + D \sum	_b m_b \dfrac{\textbf{v}_b - \textbf{v}_a}{\rho_b} \label{eq:ns-drag-sph-v} \\
\left\langle\dfrac{d e_a}{d t}\right\rangle=
 0.5\sum_b m_b \textbf{v}_{a b}\left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \cdot \nabla_a w_{a b}\left(h\right) + \sum_b \dfrac{m_b}{\rho_b}\left(\kappa_a + \kappa_b\right) \dfrac{\left(T_a - T_b\right)}{\textbf{r}_a - \textbf{r}_b} w_{ab}\left(h\right) \label{eq:ns-conduction-sph-e}
\end{align}
where the source term $\dot{\rho}$ can be a "sink" of erupted vapor due to its phase change.
%(see microphysics modelling in ATHAM \citep{oberhuber1998volcanic}).
$D$ is a drag force coefficient. $\kappa$ is the heat conduction coefficient. $T$ is the temperature. Other physics can be added easily in a similar way. Adding of these new terms leads to modification of only a few lines in the source code. The drag force term should show up when dynamic disequilibrium between different phases is considered. In that case, each phase needs one set of governing equations of Navier-Stokes type. Adding of new phase into SPH code only needs adding few new lines for the new phase besides interaction terms introduced by the new phase.

@@

\subsection{Tensile instability and corrected derivatives}
The classical SPH method was known to suffer from tensile instability and boundary deficiency. Tests of the standard SPH method indicate an instability in the tensile regime, while the calculations are stable in compression.  A simple example of such a test calculation exhibiting the instability involves a body which is subjected to an uniform initial stress, either compressive or tensile. If the initial stress is tensile, a very small velocity perturbation on a single particle can lead to particles clumping together, forming large voids and seriously corrupting density distribution. But if the initial stress is compressive, the small velocity perturbation on a single particle can not lead to any changes in particle distribution \citep{swegle1995smoothed}. To address these difficulties, \citep{chen1999improvement} proposed a corrected SPH formulation. For 1D case, employing a Taylor expansion for $A\left(x\right)$ about $x_a$, multiplying both sides by kernel function and then doing integration over the domain gives
\begin{equation}
\int_{\Omega} A\left(x\right) w\left(x- x_a, h\right) dx = 
A_a \int_{\Omega} w\left(x - x_a, h\right) dx +\dfrac{\partial A}{\partial x}(x_a) \int_{\Omega} \left(x-x_a\right) w\left(x - x_a, h\right) dx +...
\end{equation}
Ignoring derivative terms higher than first order, and writing the integral in particle approximation form leads to:
\begin{equation}
A_a = \frac{\sum_b m_b \dfrac{A_b}{\rho_b} w\left(x_a-x_b, h\right)}{\sum_b m_b \dfrac{1}{\rho_b} w\left(x_a-x_b, h\right)}
\label{eq:CSP-function-approximation-1d}
\end{equation}
Notice that the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) is actually summation approximation of Eq. (\ref{eq:SPH-kernel-normalization-prop}). That is to say, Eq. (\ref{eq:CSP-function-approximation-1d}) and Eq. (\ref{eq:SPH-approximation-sum}) are the same for particles far away from boundaries as the denominator in Eq. (\ref{eq:CSP-function-approximation-1d}) becomes $1$ in that case.
The first order derivative term can be obtained in a similar way:
\begin{equation}
\nabla A_a = \frac{\sum_b m_b \dfrac{A_b - A_a}{\rho_b} \nabla_a w\left(x_a-x_b, h\right)}{\sum_b m_b \dfrac{x_b - x_a}{\rho_b} \nabla_a w\left(x_a-x_b, h\right)}
\end{equation}

@@

\subsection{Boundary conditions}
All boundary conditions are imposed by ghost particles. Subfigure $b$ of Fig. \ref{fig:bc_and_domain_decomp} shows how boundaries are deployed. 
%Physical quantities of particles, including real particles and different types of ghost particles, are updated differently.
\subsubsection{Wall boundary condition}
Traditionally either ghost particles that mirror real particles across the boundary \citep {ferrari2009new} or boundary forces \citep {monaghan2009sph} have been used to impose the wall boundary conditions. One disadvantage of the latter  is that the boundary forces tend to corrupt the solution in the local neighborhood. In addition, a natural way of imposing eruption boundary condition is using eruption ghost particles. To impose boundary conditions in a consistent way, we adopt a modified version of the ghost particle method \citep {kumar2013parallel} for wall boundary conditions. Stationary wall ghost particles are deployed in the same way as real particles. Instead of enforcing symmetry particle by particle, a symmetric field across the boundary is explicitly enforced. Ghost particles are reflected into the domain and physical quantities are calculated at these reflected positions by SPH interpolations. It should be noted that wall ghost particles should not be counted when computing physical properties of these reflected positions. Assign all properties, except for velocity, at the corresponding reflected position to the ghost particle. The velocity of each wall ghost particles is set to have the same value but opposite direction of the interpolated velocity at its corresponding reflection. By this way, the no-slip wall boundary condition (Eq. (\ref{eq:wall_bc_v})) is imposed naturally. These wall ghost particles serve as neighbors in momentum and energy update. More implementation details about this method can be found in \citep {kumar2013parallel}. As these wall ghost particles are stationary, there is no mass flux on the boundary (Eq. (\ref{eq:wall_bc_rho}) and (\ref{eq:wall_bc_xi})). In addition, as temperature is also symmetric with respect to the boundary, the gradient of temperature vanishes and hence there is no internal energy flux on the wall boundary (Eq. (\ref{eq:wall_bc_e})). 
In our current model, the ground is assumed to be  flat. For more complicated topography, it has been shown in other work \citep {kumar2013parallel} that it also works as well. We do have some concern regarding potential limitation of this method of deployment of ghost particles for more complicated boundaries in three dimensions but that is one that is not related to the plume simulations that are the target here.

\subsubsection{Eruption boundary condition}
A natural way of imposing eruption boundary condition is using ghost particles that move with the eruption velocity and bear the temperature of the erupted material. A parabolic velocity profile that represents a fully developed Hagen-Poiseuille flow is used to determine the inlet particle velocity. The detailed shape of the parabolic profile is determined based on an averaged eruption velocity (Eq. (\ref{eq:erupt_bc_v})). The mass of eruption ghost particles is set to a value so that evaluation of Eq. (\ref{eq:ns-sph-d}) can return us a density that is consistent with the value given in Eq. (\ref{eq:erupt_bc_rho}). 
The internal energy associated with these particles are set to a value so that Eq. (\ref{eq:erupt_bc_e}) is satisfied. The mass fraction of erupted material (Eq. (\ref{eq:erupt_bc_xi})) is automatically satisfied as all particles in the eruption conduit are of phase 2. The density, momentum, and internal energy of these eruption ghost particles are not updated before they move above ground. As soon as they move out from eruption conduit, these ghost particles will be shifted to real particles and their physical quantities and position will be updated based on discretized governing equations. New ghost particles need to be added at the bottom of the eruption conduit as these existing ghost particles move upwards.

\subsubsection{Pressure boundary condition}
Another boundary condition in our model is the pressure outlet boundary. For flow in a straight channel, it is possible to treat the exit the same as the entry with a prescribed velocity profile. For flow with more complex channel, an exit far downstream of the flow disturbance is also feasible. However, the natural boundary condition (Eq. (\ref{eq:pressure_bc_p})) is more suitable for plume simulation as the outlet is open atmosphere. The way we impose pressure boundary condition is: adding several layers of pressure ghost particles surrounding the real atmosphere particles. Pressure, density and temperature are determined based on the elevation of particles. Velocity is set to zero for static atmosphere. The physical quantities for pressure ghost particles are not updated while these for real particles are updated at every time step. As position of all pressure ghost particles keeps constant, we essentially impose a static pressure boundary condition. Real particles are removed as soon as they move out of the pressure boundary.
